---
title: "NFL Analysis"
format: html
categories: [sports, football, analysis]
date: "2026-02-3"
---

How does change in ownership of an NFL impact the team?

If we take every team and look at how many wins they had each season, we can see how well each team did each season.

```{r}
#| include: false
library(nflfastR)
library(nflreadr)
library(tidyverse)
library(lubridate)
library(here)

team_df <- read_csv(here("Data/nfl_teams.csv"))
games_df <- read_csv(here("Data/nfl_games.csv")) |> mutate(home_team == case_when(home_team == "STL" ~ "LA",
                                                                                  TRUE ~ home_team)) |>
  mutate(away_team = case_when(away_team == "STL" ~ "LA",
                               TRUE ~ away_team))
```

```{r crate team wins by year}
#| include: false
historical_wins_df <- games_df |>
   select(season, home_team, away_team, home_score, away_score) |>
  mutate(winner = case_when(home_score > away_score ~ home_team,
                            home_score < away_score ~ away_team,
                            home_score == away_score ~ "tie")
  )

season_wins <- historical_wins_df |> group_by(winner, season) |> summarise(wins = n()) |> filter(winner != "tie") |> rename(team = winner)
```

First, lets look at how many times teams have won

```{r}
#| echo: false
ggplot(season_wins, aes(x = season, y = wins, group = team)) +
  geom_line(alpha = 0.2) +
  theme_minimal()
```

This isn't a very helpful visualization on its own but I want to have this here as multiple later visuals are based on this.

Let's take this data and look at it by team using a violin plot.

```{r}
#| echo: false
ggplot(season_wins, aes(x = team, y = wins)) +
  geom_violin() +
  theme_minimal() +
  coord_flip()
```

You see little some teams have clumps lower than each other but we need to see this information through time to be able to see if change in ownership was beneficial

```{r}
#| include: false

sales_df <- tibble(team = c("WAS", "DEN", "CAR", "BUF", "CLE", "JAX", "LA", "MIA"), sale_year = c(2023, 2022, 2018, 2014, 2012, 2011, 2010, 2008))

team_sales_df <- left_join(season_wins, sales_df, by = "team") |>
  mutate(sale_year = case_when(sale_year == season ~ sale_year,
                              sale_year != season ~ NA))


primary_color_df <- teams_colors_logos |> select(team_abbr, team_color) |>
  rename(team = team_abbr) |>
  mutate(team_color = case_when(
      team == "BUF" ~ "#C60C30",  # Bills bright red
      team == "CAR" ~ "#0085CA",  # Panthers blue
      team == "CLE" ~ "#311D00",  # Browns dark brown
      team == "DEN" ~ "#FF8200",  # Broncos vivid orange
      team == "JAX" ~ "#D7A22A",  # Jags gold accent
      team == "LA"  ~ "#FFD100",  # Rams yellow
      team == "MIA" ~ "#008E97",  # Dolphins aqua
      team == "WAS" ~ "#773141",  # Commanders burgundy
      TRUE             ~ team
    ))

team_sales_record_df <- team_sales_df |> left_join(primary_color_df, by = "team")

# teams that have recently sold, record for each season, and team color
recently_sold_history <- left_join(season_wins, sales_df, by = "team") |>
  filter(is.na(sale_year) == FALSE) |>
  filter(season >= sale_year) |> select(1:3) |>
  left_join(primary_color_df, by = "team") |>
  mutate(team_color = case_when(
      team == "BUF" ~ "#C60C30",  # Bills bright red
      team == "CAR" ~ "#0085CA",  # Panthers blue
      team == "CLE" ~ "#311D00",  # Browns dark brown
      team == "DEN" ~ "#FF8200",  # Broncos vivid orange
      team == "JAX" ~ "#D7A22A",  # Jags gold accent
      team == "LA"  ~ "#FFD100",  # Rams yellow
      team == "MIA" ~ "#008E97",  # Dolphins aqua
      team == "WAS" ~ "#773141"  # Commanders burgundy
    )) |>
  filter(season != 2026)

```

To see this, lets take the first plot and highlight each team after they were sold. Lets also use team colors instead of a sequential or diverging theme. This is much more helpful than before. You can see the LA Rams have been on a steep upword trend after their acquisition, contrasting this, the Miami Dolphins haven't been doing great since their sale in 2008. But how can we see if these teams have been improving since selling?

```{r plot: team recor after change in ownership}
#| echo: false
#| warning: false
ggplot(team_sales_record_df, aes(x = season, y = wins, group = team)) +
  
  # line for every team record
  geom_line(alpha = 0.2) +
  
  # line for games since sale
  geom_line(
    data = recently_sold_history,
    aes(x = season, y = wins, color = team_color),
    linewidth = 2
  ) +
  
  # point marking sale
  geom_point(
    aes(x = sale_year, y = wins, color = team_color),
    size = 4
  ) +
  
  # color by cell values
  scale_color_identity(guide = "legend",
    breaks = unique(recently_sold_history$team_color),
    labels = unique(recently_sold_history$team)) +
  
  theme_minimal() +
  labs(
    caption = "Dots mark year sold",
    title = "Team record after change in ownership",
    color = "Team Color" 
    )
```


IS SELLING A TEAM GOOD FOR THEIR RECORD

```{r building before vs after}
#| include: false

# team record before sale
old_records_df <- team_sales_record_df |> filter(team %in% c("BUF", "CAR", "CLE", "DEN", "JAX", "LA", "MIA", "WAS")) |>
  filter(case_when(team == "WAS" ~ season <= 2023,
                   team == "BUF" ~ season <= 2014,
                   team == "CAR" ~ season <= 2018,
                   team == "CLE" ~ season <= 2012,
                   team == "DEN" ~ season <= 2022,
                   team == "JAX" ~ season <= 2011,
                   team == "LA"  ~ season <= 2010, 
                   team == "MIA" ~ season <= 2008
                   ))

# team records since sale
new_records_df <- team_sales_record_df |> filter(team %in% c("BUF", "CAR", "CLE", "DEN", "JAX", "LA", "MIA", "WAS")) |>
  filter(case_when(team == "WAS" ~ season >= 2023,
                   team == "BUF" ~ season >= 2014,
                   team == "CAR" ~ season >= 2018,
                   team == "CLE" ~ season >= 2012,
                   team == "DEN" ~ season >= 2022,
                   team == "JAX" ~ season >= 2011,
                   team == "LA"  ~ season >= 2010, 
                   team == "MIA" ~ season >= 2008
                   ))
```

Lets also add linear smoothers for before sale (green), and after sale (red), to see if teams are improving.

```{r plot: team record before and after change in ownership}
#| echo: false
#| warning: false
ggplot(team_sales_record_df, aes(x = season, y = wins, group = team)) +
  
  # line showing teams since sale
  geom_line(
    data = recently_sold_history,
    aes(x = season, y = wins, color = team_color),
    linewidth = 2
  ) +
  
  # point marking sale
  geom_point(
    aes(x = sale_year, y = wins, color = team_color),
    size = 4
  ) +
  
  # line for marking before sale
  geom_line(
    data = old_records_df,
    aes(x = season, y = wins, color = team_color),
    linewidth = 2.75,
    alpha = 0.2) +
  
  # before
  geom_smooth(data = old_records_df,
              method = lm,
              se = TRUE,
              color = "green3",
              linewidth = 2.5,
              aes(group = NULL)) + 
  # after
  geom_smooth(data = new_records_df,
              method = lm,
              se = TRUE,
              color = "red",
              linewidth = 2.5,
              aes(group = NULL)) +
  
  # use colors in a cell
  scale_color_identity(
    guide = "legend",
    breaks = unique(recently_sold_history$team_color),
    labels = unique(recently_sold_history$team)
    ) +
  
  theme_minimal() +
  labs(
    caption = "Dots mark year sold. red smoother: after sale, green smoother: before sale",
    title = "Team record before and after change in ownership",
    color = "Team Color"
  )
```

We can see clearly now that teams after selling have an upward trend whereas before selling, they tended to be stagnant.

Now, lets look at if each team improved since selling, or have some struggled?

```{r}
#| echo: false
#| warning: false
ggplot(old_records_df, aes(x = season, y = wins)) +
  
  # lines for games before sale
  geom_line(aes(color = team_color), 
            linewidth = 2,
            alpha = 0.1) +
  
  # lines to make the old lines darker
  geom_line(color = "black",
            linewidth = 2,
            alpha = 0.2) +
  
  # lines for games after sale
  geom_line(data = new_records_df,
            aes(color = team_color),
            linewidth = 1.5,
            alpha = 0.75) +
  
  # lines for shading new lines
  geom_line(data = new_records_df,
            color = "black",
            linewidth = 1.5,
            alpha = 0.2) +
  
  # point to mark sale
  geom_point(
    aes(x = sale_year, y = wins, color = team_color),
    size = 4
  ) +
  
  # lm smoothers to show if team has been more successful
  # before
  geom_smooth(method = lm,
              se = FALSE,
              color = "green3",
              linewidth = 2.5) + 
  # after
  geom_smooth(data = new_records_df,
              method = lm ,
              se = FALSE, 
              color = "red",
              linewidth = 2.5) +
  
  facet_wrap(~ team, ncol = 2, scales = "free") +
  theme_void() +
  scale_color_identity(
    breaks = unique(old_records_df$team_color),
    labels = unique(old_records_df$team)
  )
```

Now we can see for each team that has sold in recent NFL history what teams have improved since selling. There is a clear trend that almost every team has a much steeper upward trajectory after sale. Why could this be? My speculation is owners sell the team when they lose interest. A new owner means renewed interest in the team financially and a general committment to better the team.
